\textit{Supplementary materials for:} Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2021-10-20 paper/main.tex
working on supplement figure captions

\section*{Results}
Before testing our main hypothesis we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task; a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendancies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

A large number of fitness-related measures displayed reliable
correlations (for a complete report, see Fig.~\fitFitCorr).  For
example, body mass index (BMI) and weight were correlated
($r = 0.91, p < 0.0001$).  Resting heart rate over
the prior week was negatively correlated with recent
low-to-moderate-intensity (``fat burn'') cardiovascular
activity levels ($r = 0.70, p = 0.0004$).
Participants' peak heart rate (averaged over the prior week) were also
negatively correlated with recent increases in step counts and daily
elevation gains ($r\mathrm{s} < -0.26, p\mathrm{s} < 0.03$, boostrap
corrected), where recent changes were defined as the average values
over the seven days leading up to the test day divided by the average
values over the preceding 30 days.  Although several demographic
attributes (Fig.~\demDemCorr) displayed trivial correlations (e.g.,
participants identifying as male never reported identifying as female,
and so on), we also observed a negative correlation between reported
stress and alertness ($r = -0.44, p < 0.0001$), and
positive correlations between the reported clarity of the instructions for all
tasks ($r\mathrm{s} > 0.26, p\mathrm{s} < 0.02$).

@@

The above analyses indicate that recent differences in fitness-related
activity are associated with differences in memory performance and
mental health measures.  Although the analyses treated these measures
on average or in aggregate, many of the measures we collected are
dynamic. For example, the amount or intensity of physical exercise
people engage in can vary over time, and so on.  We wondered whether
the dynamics of fitness-related measures might related to memory
performance and mental health measures.  To this end, we carried out a
series of reverse correlation analyses (see \textit{Reverse
  correlation analyses}) to examine whether participants with
different cognitive or mental health profiles also tended to display
differences in fitness-related measures over time.  In particular, we
examined fitness data collected from participants' Fitbit devices over
the year prior to their test day in our study.  Several example findings are summarized in
Figure~\ref{fig:dynamics}.  We found that participants who performed
well on the immediate and delayed free recall memory tests and on the
naturalistic recall tests tended to be more active than participants
who performed poorly on those tests (Figs.~\ref{fig:dynamics}a, b;
\activityTimecourse).  Conversely, participants who performed well on
the immediate and delayed foreign language flashcards tasks tended to
be \textit{less} active.  These differences were present even a full
year before the testing day.  We also found substantial variability
across people with different (self-reported) mental health profiles
(Figs.~\ref{fig:dynamics}c, \activityTimecourseMH).  Due to small sample
sizes of individuals exhibiting several mental health dimensions, it
is difficult to distinguish generalizable trends from individual
differences that one or two individuals happened to exhibit.  However,
several large-sample-size trends emerged.  For example, participants
who reported higher levels of stress also tended to be slightly more
physically active than participants who reported lower stress levels.
We found analogous differences in other activity-related measures
(Figs.~\activityTimecourse~and \activityTimecourseMH), cardiovascular
measures (Figs.~\cardioTimecourse~and \cardioTimecourseMH), and
sleep-related measures (Figs.~\sleepTimecourse~and
\sleepTimecourseMH).  Taken together, the analyses suggest that
cogntive and mental health differences are also associated with differences
in the dynamics of physical health measures.

\section*{Discussion}

  \begin{itemize}
  \item summarize key findings
  \item correlation versus causation
         \item what can vs. can't we know?  we can identify correlations, but not causal direction-- e.g. we cannot know whether exercise \textit{causes} mental changes versus whether people with particular neural profiles might tend to engage in particular exercise behaviors.  that being said, we \textit{can} separate out baseline tendencies (e.g., how people tend to exercise in general) versus recent changes (e.g., how they happened to have exercised prior to the experiment).
  \item related work (exercise/memory, exercise/mental health), what this study adds
    \item future direction: towards customized physical exercise recommendation engine for optimizing mental health and mental fitness
    \end{itemize}

>>>>

\section*{Results}
Before testing our main hypothesis we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task; a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

A large number of fitness-related measures displayed reliable
correlations (for a complete report, see Fig.~\fitFitCorr).  For
example, body mass index (BMI) and weight were correlated
($r = 0.91, p < 0.0001$).  Resting heart rate over
the prior week was negatively correlated with recent
low-to-moderate-intensity (``fat burn'') cardiovascular
activity levels ($r = 0.70, p = 0.0004$).
Participants' peak heart rate (averaged over the prior week) were also
negatively correlated with recent increases in step counts and daily
elevation gains ($r\mathrm{s} < -0.26, p\mathrm{s} < 0.03$, bootstrap
corrected), where recent changes were defined as the average values
over the seven days leading up to the test day divided by the average
values over the preceding 30 days.  Although several demographic
attributes (Fig.~\demDemCorr) displayed trivial correlations (e.g.,
participants identifying as male never reported identifying as female,
and so on), we also observed a negative correlation between reported
stress and alertness ($r = -0.44, p < 0.0001$), and
positive correlations between the reported clarity of the instructions for all
tasks ($r\mathrm{s} > 0.26, p\mathrm{s} < 0.02$).

@@

The above analyses indicate that recent differences in fitness-related
activity are associated with differences in memory performance and
mental health measures.  Although the analyses treated these measures
on average or in aggregate, many of the measures we collected are
dynamic. For example, the amount or intensity of physical exercise
people engage in can vary over time, and so on.  We wondered whether
the dynamics of fitness-related measures might related to memory
performance and mental health measures.  To this end, we carried out a
series of reverse correlation analyses (see \textit{Reverse
  correlation analyses}) to examine whether participants with
different cognitive or mental health profiles also tended to display
differences in fitness-related measures over time.  In particular, we
examined fitness data collected from participants' Fitbit devices over
the year prior to their test day in our study.  Several example findings are summarized in
Figure~\ref{fig:dynamics}.  We found that participants who performed
well on the immediate and delayed free recall memory tests and on the
naturalistic recall tests tended to be more active than participants
who performed poorly on those tests (Figs.~\ref{fig:dynamics}a, b;
\activityTimecourse).  Conversely, participants who performed well on
the immediate and delayed foreign language flashcards tasks tended to
be \textit{less} active.  These differences were present even a full
year before the testing day.  We also found substantial variability
across people with different (self-reported) mental health profiles
(Figs.~\ref{fig:dynamics}c, \activityTimecourseMH).  Due to small sample
sizes of individuals exhibiting several mental health dimensions, it
is difficult to distinguish generalizable trends from individual
differences that one or two individuals happened to exhibit.  However,
several large-sample-size trends emerged.  For example, participants
who reported higher levels of stress also tended to be slightly more
physically active than participants who reported lower stress levels.
We found analogous differences in other activity-related measures
(Figs.~\activityTimecourse~and \activityTimecourseMH), cardiovascular
measures (Figs.~\cardioTimecourse~and \cardioTimecourseMH), and
sleep-related measures (Figs.~\sleepTimecourse~and
\sleepTimecourseMH).  Taken together, the analyses suggest that
cognitive and mental health differences are also associated with differences
in the dynamics of physical health measures.

\section*{Discussion}
After collecting a year's worth of fitness-tracking data from 113
participants, we ran each participant in a battery of memory tasks and
had them fill out a series of demographic and mental health-related
questions.  We found that the associations between fitness-related
activities, memory performance, and mental health are heterogeneous.
Our results suggest that engagement in particular physical
activities (e.g., differing in time relative to the
test day, intensity, duration, etc.) is also reflected in participants'
memory performance patterns across tasks and in participants' mental
health attributes.

One important limitation of our study is that we cannot distinguish
correlations between different measures from potential causal
effects.  For example, we cannot know (from our study) whether
engaging in particular forms of exercise \textit{causes} changes in
memory performance or mental health, or whether (alternatively) people
who tend to engage in similar forms of exercise also happen to
exhibit similar memory and/or mental health profiles.  In other words,
an overlapping set of processes or person-specific attributes may lead
someone to both form particular habits around exercise and display high or low performance on a
given memory test.  We do not know whether memory performance or
aspects of mental health might be manipulated or influenced by
changing the patterns of physical activity someone engages in.  For
this reason, we have been careful to frame our findings as
correlations and associations, rather than to imply knowledge about a causal direction
of our findings.

Although the present study cannot reveal causal effects, a large prior
literature provided some insight into potential causal effects by examining the neural and cognitive effects of a variety
of exercise interventions~\citep{ChanEtal15, VidoEtal15, KamiEtal07,
  ImboEtal19, SuwaEtal17, SinhEtal21, Tomp03}.  A limitation of that
prior work is that most of these studies examine how relatively
short-term changes in exercise (e.g., on timescales of hours to days
or, rarely, weeks to months) affect a cognitive performance on single
task or aspect of mental health.  The present study examines
longer-term exercise (over a full year), and relates long-term
exercise history to performance on a variety of tasks and to a variety
of mental health dimensions.

To the extent that exercise \textit{does} provide a non-invasive means
of manipulating cognitive performance and mental health, our work may
have exciting implications for cognitive enhancement.  For example,
one might imagine building a recommendation system that suggests a
particular exercise regimen tailored to improve a specific aspect of
an individual's cognitive performance (e.g., the efficacy of a student's study session for an
upcoming exam) or mental health (e.g., reducing symptoms of anxiety
before an important meeting).  Just as strength training may be
customized to target a specific muscle group, or to improve
performance on a specific physical task, similar principles might also
be applied to target specific improvements in cognitive fitness
and mental health.

### 2021-10-20 paper/supplement.tex
working on supplement figure captions

@@

\begin{sidewaysfigure}[p]
\centering
\includegraphics[width=1\textwidth]{figs/free_recall_behavior}
\caption{\textbf{Free recall behavioral results.}}
\label{fig:fr_behavioral}
\end{sidewaysfigure}

\begin{sidewaysfigure}[p]
\centering
\includegraphics[width=1\textwidth]{figs/naturalistic_recall_behavior}
\caption{\textbf{Naturalistic recall behavioral results.}}
\label{fig:nat_behavioral}
\end{sidewaysfigure}

\begin{figure}[p]
\centering
\includegraphics[width=0.6\textwidth]{figs/vocab_learning_behavior}
\caption{\textbf{Foreign language vocabulary learning behavioral results.}}
\label{fig:vocab_behavioral}
\end{figure}

@@

\bibliographystyle{apa}
\bibliography{/Users/jmanning/CDL-bibliography/cdl}
\end{document}

>>>>

\newcommand{\task}{1}
\newcommand{\immediateBehavior}{2}
\newcommand{\delayedBehavior}{3}
\newcommand{\fitness}{4}
\newcommand{\MHcorr}{5}
\newcommand{\dynamics}{6}

@@

\begin{sidewaysfigure}[p]
\centering
\includegraphics[width=1\textwidth]{figs/free_recall_behavior}
\caption{\textbf{Free recall behavioral results.  a--e. Immediate free
  recall.}  \textbf{a. Probability of first recall.}  Probability of
recalling each studied word first, as a function of its presentation
position.  \textbf{b. Lag Conditional Response Probability.}
Probability of recalling the word presented at position $i +
\mathrm{Lag}$ following the recall of the word presented at position
$i$.  \textbf{c. Clustering scores.}  Each score denotes participants'
tendancies to successively recall (cluster) words according to the
given feature dimension~\citep{PolyEtal09}: word length, starting letter, (semantic)
category, size (large or small), or presentation position (temporal).
\textbf{d. Serial position curve.}  Probability of recalling each word
as a function of its presentation position.  \textbf{e. Recall
  accuracy.}  Distribution of the average proportions of recalled
words, across all lists studied by each participant.  \textbf{f--j.
  Delayed free recall.}  These panels are in the same formats as
Panels a -- e, but they reflect performance on the delayed free recall
memory tests.  All panels: error bars and error ribbons denote bootstrap-estimated
95\% confidence intervals.  Shading (saturation) denotes results for
different subsets of participants, according to the average
proporition of words they remembered (group boundaries are indicated
by the colored lines in Panels e and j).}
\label{fig:fr_behavioral}
\end{sidewaysfigure}

\begin{sidewaysfigure}[p]
\centering
\includegraphics[width=1\textwidth]{figs/naturalistic_recall_behavior}
\caption{\textbf{Naturalistic recall behavioral results.}
  \textbf{a. Identified topics.}  Applying a topic model to a series
  of sliding windows of the video's transcript~\citep{HeusEtal21}
  revealed a set of 13 unique (non-trivial) topics.  The Panel
  displays the 5 top-weighted words from each topic.  \textbf{b. Topic
  timecourse of the video.}  Each row displays the topic weights for a
single moment of the video.  \textbf{c. Topic correlation matrix.}  The
correlations between the topic vectors for each pair of moments from
the video reveals an event-like block diagonal structure.
\textbf{d. Video topic trajectory.}  The topic video's topic
timecourse (Panel b) has been projected onto 2-dimensions using
Uniform Manifold Approximation and
Projection~\citep[UMAP;][]{McInEtal18}.  Each colored dot reflects an
event, identified by applying a Hidden Markov Model to the video's
topic timecourse~\citep{BaldEtal17, HeusEtal21}.  Red dots denote
earlier timepoints in the video and blue dots denote later
timepoints.  \textbf{e. Immediate recall trajectory.}  The black curve
displays the average
topic timecourse (projected into 2D using UMAP), obtained by applying the topic model shown in Panel A
to the participants' written transcripts from the immediate recall
test.  The arrows denote agreement across participants in the
directions of their topic trajectories, for participants whose
trajectories intersected the corresponding region of topic space.
Blue arrows denote reliable agreement across participants ($p < 0.05$, corrected).
\textbf{f. Delayed recall trajectory.} This panel is in the same
format as Panel e, but displays the trajectory for participants'
delayed recall of the video.  \textbf{g. Immediate recall precision.}  Distribution of
average recall \textit{precision}, across all of the events each
participant recalled during the immediate recall test.  Precision is
defined as the correlation between the topic vector for a given
recalled event and the best-matching (most highly correlated) video
event's topic
vector~\citep{HeusEtal21}.  \textbf{h. Delayed recall precision.}
This panel is in the same format as Panel g, but displays the average
precision values for the delayed memory test.}
\label{fig:nat_behavioral}
\end{sidewaysfigure}

\begin{figure}[p]
\centering
\includegraphics[width=0.6\textwidth]{figs/vocab_learning_behavior}
\caption{\textbf{Foreign language vocabulary learning behavioral
    results.}  \textbf{a. Immediate recall....}}
\label{fig:vocab_behavioral}
\end{figure}

@@

  \clearpage
  \newpage
\bibliographystyle{apa}
\bibliography{/Users/jmanning/CDL-bibliography/cdl}
\end{document}
